---
title: "About"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
HAS_GEO <- requireNamespace("geoscales", quietly = TRUE) &&
           requireNamespace("dplyr", quietly = TRUE)
```

`r·en·euro` provides a European energy system model built on
[energyRt](https://energyRt.org). It was developed as teaching material for a
modelling course, and as a standalone working model.

It currently reproduces
[PyPSA-Eur](https://github.com/PyPSA/pypsa-eur), an established open model of
the European power system, at several spatial resolutions. A published model
provides a reference solution for verifying the R implementation.

**Current functionality includes:**  
- Import of pre-built PyPSA-Eur models from netCDF, including component
  attributes that PyPSA omits when they hold default values.
- Translation into `energyRt` model objects: regions, commodities, demand,
  generation technologies, storage, hydro, weather and transmission.
- Sampling in time (subsets of snapshots) and in space (bus aggregation or
  sub-territory selection).
- Solution through any `energyRt` back end — GLPK, Julia/JuMP, Python/Pyomo, GAMS — and the solvers each supports, locally or remotely via NEOS.
- Comparison of converted results against the source model, by quantity.

Transmission is represented along a configurable ladder, from an unconstrained
copperplate through capacity-limited transport with piecewise-linear losses to
Kirchhoff's voltage law, since `energyRt` uses a transport formulation while
PyPSA combines both.

## Models

The release ships a range of spatial resolutions over the same European
system, all on the full year (8,760 hourly snapshots) unless stated, produced by
two routes.

**Built by PyPSA-Eur.** Clustered by PyPSA-Eur's `cluster_network` rule and read
into energyRt without further reduction, so each is comparable to a PyPSA-Eur
run at the same size.

| model | nodes | how it was clustered | period |
|---|---:|---|---|
| `pypsa_eur_5` | 5 | k-means, Belgium only | one week, 168 h |
| `pypsa_eur_5cp` | 5 | as above, copperplate | one week, 168 h |
| `pypsa_eur_41` | 41 | k-means | full year |
| `pypsa_eur_250` | 250 | k-means | full year |
| `pypsa_eur_nuts3` | 1,035 | NUTS3 busmap, **not k-means** † | full year |

† `pypsa_eur_nuts3` uses the same PyPSA-Eur workflow on a busmap generated by
this package (`data-raw/make_busmap.R`) rather than by k-means, so its regions
are administrative. Everything downstream — renewable profiles, line
aggregation, costs — is PyPSA-Eur's.

**Aggregated in R.** Derived from `pypsa_eur_nuts3` with `aggregate_pypsa()`,
which applies PyPSA's reduction strategies without a second PyPSA-Eur run. Each
takes seconds to minutes rather than a rebuild from raw data.

| model | nodes | aggregated to | build | period |
|---|---:|---|---:|---|
| `pypsa_eur_nuts0` | 36 | NUTS0, countries | 16 s | full year |
| `pypsa_eur_nuts1` | 106 | NUTS1 | 34 s | full year |
| `pypsa_eur_nuts2` | 289 | NUTS2 | 75 s | full year |

`pypsa_eur_nuts2` (289 nodes, aggregated in R) and `pypsa_eur_250` (PyPSA's
k-means) are the same system at nearly the same size by two independent methods,
so either can be used to check the other.

**NUTS** — *Nomenclature of Territorial Units for Statistics* — is Eurostat's
hierarchy of European regions, against which national statistics are published.
It has four nested levels, each a complete partition of the territory:

| level | Eurostat's definition | count | one nesting chain |
|---|---|---:|---|
| NUTS0 | country | 36 | `DE` |
| NUTS1 | major socio-economic regions | 109 | `DE9` |
| NUTS2 | basic regions for regional policy | 296 | `DE93` |
| NUTS3 | small regions for specific diagnoses | 1,477 | `DE933` |

Each code contains its ancestry: `DE933` is inside `DE93`, inside `DE9`,
inside `DE`, so the hierarchy is read rather than inferred.

The 2021 edition is used, as in PyPSA-Eur. An official classification lets a
model be joined to published statistics — population, GDP, industrial output —
without a separate crosswalk.

Four countries in the model territory are outside NUTS: Bosnia and
Herzegovina, Moldova, Ukraine and Kosovo. They are filled from OpenStreetMap
first-level administrative boundaries, so "NUTS1" and "NUTS2" for those four are
stand-ins rather than real levels.

The four NUTS models nest over the same system, so a result can be tested for
sensitivity to spatial resolution with nothing else changed. Node counts fall
short of region counts because a region without a substation merges into a
neighbour, keeping its demand and generation: 109 NUTS1 regions give 106 nodes,
296 NUTS2 give 289, and 1,477 NUTS3 give 1,035. Every country has at least one
substation, so NUTS0 is 36 for 36.

**`pypsa_eur_250`** is the recommended working model. PyPSA-Eur is normally
clustered to 50–250 nodes for computational reasons; its k-means is
load-weighted and respects country boundaries.

**`aggregate_pypsa()`** applies PyPSA's strategies: extensive quantities sum,
intensive ones take a capacity-weighted mean and so stay within their members'
range, intra-cluster branches are dropped, and corridor lengths are recomputed
from the new region centroids with impedance rescaled to follow. It does not
re-run the clustering from raw data, which is why a NUTS2 model takes 75 seconds
rather than a Snakemake run.

## Regions

`nuts_gs` is the region hierarchy the four NUTS models are built from: a
[geoscales](https://github.com/optimal2050/geoscales) object with 1,477 NUTS3
atoms nesting through NUTS2, NUTS1 and country up to a single European root.
Aggregating a model to any level, or splitting a national figure down to
regions, is a call against this object.

It carries per-region data as well as geometry, in two families. **Geographic**
— area, population, GDP per capita and GDP total — is defined for every region.
**Model** — annual demand, coincident peak, substation count, existing
generation and hydro capacity, and the wind and solar potentials by carrier —
comes from the NUTS3 network and is located at substations, so it is zero in
the 442 regions that contain none.

Seven columns are declared weights. The choice among them is a modelling
decision: national demand divides by `load_twh` or `pop`, a wind target by
`pot_onwind`; area is rarely the right weight for either.

```{r demand-wind, eval = HAS_GEO, message = FALSE}
library(reneuro)
library(dplyr)

geoscales::geoscale_leaftable(nuts_gs) |>
  as.data.frame() |>
  group_by(nuts0) |>
  summarise(twh = sum(load_twh), wind_gw = sum(pot_onwind) / 1e3) |>
  arrange(desc(twh)) |>
  head()
```

The solar and offshore potentials overlap. Fixed-tilt and tracking
photovoltaics occupy the same land, and 313 of the 314 offshore regions carry
more than one offshore type on the same sea area, so these columns are kept
separate and never summed.

## A source model

`pypsa_eur_nuts3` is not intended to be solved at full resolution. A full-year
continental model at 1,035 nodes is beyond the open solvers this package
targets: on a 96-hour sample it is 1.3 million LP rows and about 18 minutes, and
at four weeks it is 9.4 million rows, where the interior-point method returns no
solution. Use `pypsa_eur_250` or `pypsa_eur_nuts2` at continental scale.

Its use is subsetting. It holds the finest granularity the underlying data
supports, so a country or study area can be taken from it at full detail and
solved on its own. The coarser models cannot supply that detail, having already
averaged it away.

<details>
<summary>Show the code</summary>

```r
library(dplyr)

pt <- geoscales::geoscale_leaftable(nuts_gs) |>
  as.data.frame() |>
  filter(nuts0 == "PT") |>
  pull(region) |>
  intersect(get_region(pypsa_eur_nuts3))

m <- attach_weather(pypsa_eur_nuts3, c("onwind", "solar", "ror")) |>
  energyRt::subset_model_regions(region = pt)
```

</details>

The result is Portugal at 22 nodes: every NUTS3 region containing a
substation, with the network between them intact. `region` accepts any subset of
the model's regions, so a study area need not follow a national border.

[`subset_model_regions()`](https://energyRt.org/reference/subset_model_regions.html)
drops the corridors crossing the boundary and reports each. Without replacement
the model is islanded and must meet demand from its own resources, so its cost
is an upper bound; `boundary_prices` replaces the dropped routes with priced
import/export stubs.

The weather series are the bulk of the model. They ship as separate objects by
resource — `wx_nuts3_onwind`, `wx_nuts3_solar` and so on — and are attached on
demand:

```r
m <- attach_weather(pypsa_eur_nuts3)                  # all resources
m <- attach_weather(pypsa_eur_nuts3, c("onwind", "solar"))
```

Models are lazy-loaded and require neither Python nor a PyPSA-Eur clone, and
each carries its provenance as an attribute. `data-raw/` holds the scripts that
rebuild every object from a clone, recording the upstream commit and
configuration used.

## Solver benchmarks

Measured on one workstation. The same interpolated scenario was given to each
back end, so the linear programme is identical and the objectives agree to
solver tolerance in every case. Times are end to end, including model
generation and reading the solution back.

| model | rows (LP) | GLPK | CBC | HiGHS |
|---|---:|---:|---:|---:|
| 5 nodes, 168 h | ~24 k | **10 s** | 19 s | 24-47 s |
| 69 nodes, 96 h | ~300 k | 2,158 s | 809 s | **79 s** |
| 1,035 nodes, 96 h | 1.3 M | -- | -- | **18.5 min** |

The ordering reverses with size. GLPK is fastest on the smallest model, where
start-up cost dominates; at 69 regions HiGHS is 27 times faster, and the margin
grows from there. Julia adds a fixed 20-30 s per invocation for just-in-time
compilation, which is amortised across a session but not within a single call,
so Pyomo is faster end to end for one small solve even where Julia's solver call
is faster.

The NUTS3 model on a four-week sample (672 h) produces 9.4 million rows after
presolve, where the interior-point method returns no solution. First-order
(`pdlp`) and dual simplex methods have not been tested at that size.

Reproduce with [`benchmark_solvers()`](reference/benchmark_solvers.html):

```r
scen <- energyRt::interpolate_model(pypsa_eur_5, name = "be")
benchmark_solvers(scen)
```

## Development

Planned work, in order of priority:

1. Revision and extension of the input data, including alternative sources.
2. Review of the processing algorithms that derive model parameters from raw
   data: aggregation, clustering, capacity factors and loss representation.
3. Extension to sectors beyond electricity.
4. Interfaces to other modelling frameworks for cross-model comparison.

## Contributing

Contributions are welcome. Issues and pull requests may be opened at
[github.com/optimal2050/reneuro](https://github.com/optimal2050/reneuro).

The package follows the
[optimal2050 conventions](https://github.com/optimal2050/.github/blob/main/CONVENTIONS.md):
tidyverse style with the native pipe, roxygen2 with Markdown, and the shared
pkgdown theme.

## References

This package reproduces PyPSA-Eur. If you use `reneuro` in research, please
cite the upstream work as well as this package.

**The software**

- [PyPSA](https://github.com/PyPSA/pypsa) — the power system analysis
  framework PyPSA-Eur is built on.
- [PyPSA-Eur](https://github.com/PyPSA/pypsa-eur) — the European model
  reproduced here, version `v2026.02.0`.
- [energyRt](https://github.com/optimal2050/energyRt) — the modelling framework
  `reneuro` is built on.

**The papers**

For the electricity-only model, which is what this package currently
reproduces:

> Jonas Hörsch, Fabian Hofmann, David Schlachtberger and Tom Brown (2018).
> **PyPSA-Eur: An open optimisation model of the European transmission
> system.** *Energy Strategy Reviews* **22**, 207–215.
> [doi:10.1016/j.esr.2018.08.012](https://doi.org/10.1016/j.esr.2018.08.012),
> [arXiv:1806.01613](https://arxiv.org/abs/1806.01613)

For the underlying framework:

> Tom Brown, Jonas Hörsch and David Schlachtberger (2018).
> **PyPSA: Python for Power System Analysis.**
> *Journal of Open Research Software* **6**(1), 4.
> [doi:10.5334/jors.188](https://doi.org/10.5334/jors.188),
> [arXiv:1707.09913](https://arxiv.org/abs/1707.09913)

On spatial resolution and what coarsening removes:

> Jonas Hörsch and Tom Brown (2017). **The role of spatial scale in joint
> optimisations of generation and transmission for European highly renewable
> scenarios.** *14th International Conference on the European Energy Market
> (EEM)*.
> [doi:10.1109/EEM.2017.7982024](https://doi.org/10.1109/EEM.2017.7982024),
> [arXiv:1705.07617](https://arxiv.org/abs/1705.07617)

PyPSA-Eur is now a sector-coupled model, and its authors ask that sector studies
cite instead:

> Fabian Neumann, Elisabeth Zeyen, Marta Victoria and Tom Brown (2023).
> **The potential role of a hydrogen network in Europe.** *Joule* **7**, 1–25.
> [doi:10.1016/j.joule.2023.06.016](https://doi.org/10.1016/j.joule.2023.06.016)

> Marta Victoria, Elisabeth Zeyen and Tom Brown (2022). **Speed of technological
> transformations required in Europe to achieve different climate goals.**
> *Joule* **6**(5), 1066–1086.
> [doi:10.1016/j.joule.2022.04.016](https://doi.org/10.1016/j.joule.2022.04.016)

Each release of PyPSA and PyPSA-Eur carries a version-specific DOI on Zenodo,
[10.5281/zenodo.3946412](https://doi.org/10.5281/zenodo.3946412) and
[10.5281/zenodo.3520874](https://doi.org/10.5281/zenodo.3520874) respectively.
Cite the version used.

**Citing `reneuro`**

```r
citation("reneuro")
```

Machine-readable metadata is in `CITATION.cff`.

## Licence

| | |
|---|---|
| `reneuro` sources | Apache-2.0 |
| `reneuro` with `energyRt` | AGPL-3 |

`energyRt` is licensed under AGPL-3 and is imported by `reneuro`. R links
imported namespaces into a single process, so distributing the two together
constitutes a combined work conveyed under AGPL-3. Apache-2.0 is one-way
compatible with AGPL-3. The conversion code alone, independent of `energyRt`,
remains available under Apache-2.0.

Converted models inherit the licences of their inputs. OpenStreetMap supplies
the transmission topology under ODbL-1.0, whose share-alike terms extend to
derived databases; Copernicus land cover and EEA Natura 2000 data are
CC-BY-4.0; Eurostat NUTS geometry is reusable under Commission Decision
2011/833/EU, with acknowledgement of the European Union and EuroGeographics.
Two electricity demand series (OPSD and ENTSO-E) are recorded as licence
*unknown* in PyPSA-Eur's data inventory; this is unresolved upstream.

Shipped data carries a `LICENSE.note` recording provenance for each object,
including inputs excluded as non-redistributable.

