---
title: "Data in PyPSA-Eur 41-node"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data in PyPSA-Eur 41-node}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5.6, dpi = 96)
MAPS <- requireNamespace("geoscales", quietly = TRUE) &&
        requireNamespace("sf", quietly = TRUE) &&
        requireNamespace("ggplot2", quietly = TRUE)
PAL <- MAPS && requireNamespace("energypal", quietly = TRUE)
```

```{r setup, message = FALSE}
library(reneuro)
library(dplyr)
```

This article shows the *processed* data that enters the 41-node model —
regions, network, weather, demand and capacities — drawn live from the
shipped `pypsa_eur_41` and its geoscale. Everything upstream (the raw
sources, the NUTS datasets and their aggregation arithmetic) lives in the
[data-sources article](data-sources.html).

```{r objs, include = FALSE}
objs <- pypsa_eur_41@data[[1]]@data
cls <- vapply(objs, function(o) class(o)[1], "")
```

## The regions

PyPSA-Eur clusters its substation network to 41 onshore regions — most
countries are one node; the largest split into several. The polygons ship
with the model as the `bus` geoframe of `pypsa_eur_41_gs`:

```{r regions, eval = MAPS, fig.cap = "The 41 model regions (PyPSA-Eur onshore clustering). Clusters never cross country borders."}
library(ggplot2)
gs <- pypsa_eur_41_gs
geo <- geoscales::geoscale_geometry(gs, "bus")
cc <- geoscales::geoscale_geometry(gs, "country")

ggplot() +
  geom_sf(data = geo, fill = "wheat", colour = "white", linewidth = 0.4) +
  geom_sf(data = cc, fill = NA, colour = "grey35", linewidth = 0.3) +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", colour = NA))
```

## The network

Transmission enters the model as `trade` corridors between region
centroids — rated capacities from the summed thermal limits of the
underlying lines:

```{r network, eval = MAPS, fig.height = 6.5, fig.cap = "The 41-node network: buses and trade corridors, line width by rated capacity."}
ctr <- suppressWarnings(sf::st_centroid(geo))
xy <- as.data.frame(sf::st_coordinates(ctr))
xy$bus <- geo$bus

tr <- objs[cls == "trade"]
routes <- unique(do.call(rbind, lapply(tr, function(t) {
  mw <- t@capacity$stock[1]
  if (is.null(mw) || is.na(mw)) mw <- t@capacity$cap.lo[1]
  cbind(t@routes[1, , drop = FALSE], MW = mw)
})))
seg <- routes |>
  left_join(xy, by = c(src = "bus")) |>
  rename(x0 = X, y0 = Y) |>
  left_join(xy, by = c(dst = "bus")) |>
  rename(x1 = X, y1 = Y)

ggplot() +
  geom_sf(data = geo, fill = "wheat", colour = "white", linewidth = 0.3) +
  geom_segment(data = seg, aes(x0, y0, xend = x1, yend = y1,
                               linewidth = MW / 1e3),
               colour = "dodgerblue", alpha = 0.8) +
  geom_point(data = xy, aes(X, Y), size = 0.7, colour = "grey25") +
  scale_linewidth(range = c(0.2, 2.2)) +
  labs(linewidth = "GW") +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", colour = NA))
```

## Weather: capacity factors per region

The model's VRE profiles are per-region hourly capacity factors (the
`weather` objects, 8,760 hours of 2025). Their annual means show what
each region's blended resource offers:

```{r cf-helper, include = FALSE, eval = MAPS}
cf_map <- function(wname) {
  w <- objs[[wname]]@weather |>
    summarise(cf = mean(wval), .by = region)
  geo |> left_join(w, by = c(bus = "region"))
}
```

```{r cf-wind, eval = PAL, fig.cap = "Mean onshore-wind capacity factor of the blended regional profile (`W_ONWIND`), 2025."}
ggplot(cf_map("W_ONWIND")) +
  geom_sf(aes(fill = cf), colour = "white", linewidth = 0.3) +
  energypal::scale_fill_energy_c("windatlas", name = "mean CF") +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", colour = NA))
```

```{r cf-solar, eval = PAL, fig.cap = "Mean solar capacity factor of the blended regional profile (`W_SOLAR`), 2025."}
ggplot(cf_map("W_SOLAR")) +
  geom_sf(aes(fill = cf), colour = "white", linewidth = 0.3) +
  energypal::scale_fill_energy_c("solaratlas", name = "mean CF") +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", colour = NA))
```

The blending is optimistic by construction — every region offers its
whole eligible area at close to its best-sites capacity factor.

## Demand

Measured 2025 load, distributed to regions by the GDP/population proxy:

```{r demand, eval = MAPS, fig.cap = "Annual demand per region, TWh (measured ENTSO-E 2025, distributed below country level by proxy)."}
dem <- bind_rows(lapply(objs[cls == "demand"], function(d) d@demand)) |>
  summarise(TWh = sum(demand) / 1e6, .by = region)
ggplot(geo |> left_join(dem, by = c(bus = "region"))) +
  geom_sf(aes(fill = TWh), colour = "white", linewidth = 0.3) +
  scale_fill_viridis_c(option = "C", trans = "sqrt") +
  theme_void() +
  theme(plot.background = element_rect(fill = "white", colour = NA))
```

```{r demand-total}
sum(dem$TWh)
```

## Capacities and potentials

Per carrier: the existing fleet (`stock` for the non-extendable, `cap.lo`
floors for the extendable), and the buildable potential (`cap.up`):

```{r capacities}
cap <- bind_rows(lapply(objs[cls == "technology"], function(t) {
  d <- t@capacity
  tibble(tech = t@name,
         stock_GW = sum(d$stock, na.rm = TRUE) / 1e3,
         floor_GW = sum(d$cap.lo, na.rm = TRUE) / 1e3,
         potential_GW = sum(d$cap.up, na.rm = TRUE) / 1e3)
}))
cap |>
  filter(stock_GW + floor_GW + potential_GW > 0) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  arrange(desc(pmax(stock_GW, floor_GW)))
```

Potentials are flat area sums (eligible km² × deployment density) — the
[data-sources article](data-sources.html) documents how they and
the profiles are made, and the known caveats.

## See also

* `vignette("reneuro")` — a model end to end
* [data-sources](data-sources.html) — the upstream sources and
  the NUTS datasets with their aggregation arithmetic
* `vignette("about")` — how each model was built
