---
title: "The data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5.2, dpi = 96)
HAS_GEO <- requireNamespace("geoscales", quietly = TRUE) &&
           requireNamespace("sf", quietly = TRUE)
HAS_GG  <- requireNamespace("ggplot2", quietly = TRUE)
MAPS    <- HAS_GEO && HAS_GG

# `fold = TRUE` collapses a chunk's SOURCE behind a disclosure triangle and
# leaves its output alone, so the plumbing chunks stop interrupting the prose
# while their tables and maps stay on the page.
`%||%` <- function(x, y) if (is.null(x)) y else x
local({
  hook <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    out <- hook(x, options)
    if (!isTRUE(options$fold)) return(out)
    label <- options$fold_label %||% "Show code"
    paste0("<details>\n<summary>", label, "</summary>\n\n",
           out, "\n\n</details>\n")
  })
})
```

```{r setup, message = FALSE}
library(reneuro)
library(dplyr)
```

This article uses only shipped data. It needs no PyPSA-Eur clone and no solver.

Three datasets describe the regions the models are built on: `nuts_gs` (the
geography and the data attached to it), `nuts_load` (demand) and `nuts_lines`
(transmission). Each section below aggregates one of them from NUTS3 to coarser
levels and gives the arithmetic used, which differs by quantity: capacities sum,
per-capita values take a weighted mean, and peaks and impedances have to be
recomputed rather than combined.

## 1. Geography: five nested levels

```{r frames, eval = HAS_GEO}
gs <- nuts_gs
vapply(geoscales::geoscale_geoframes(gs),
       function(f) length(geoscales::geoscale_regions(gs, f)), 0L)
```

1,477 NUTS3 atoms group into 296 NUTS2, 109 NUTS1, 36 countries and a single
`europe` root. Every finer code has exactly one parent; the build asserts this
rather than assuming it, and the aggregation below depends on it.

The `europe` root is the only region at which a continental commodity balance
can be declared. `newCommodity(geoframe = "europe")` gives one Europe-wide CO2
allowance or gas market; without it the coarsest balance is per country, i.e. 36
separate markets.

```{r stack, eval = MAPS, fig.height = 6}
energyRt::plot_geoscale(gs, type = "stack")
```

The same hierarchy as an icicle plot:

```{r icicle, eval = MAPS, fig.height = 4}
energyRt::plot_geoscale(gs, type = "icicle") +
  ggplot2::labs(title = NULL)
```

Four countries — **BA, MD, UA, XK** — have no NUTS at all and are filled from
OpenStreetMap `adm1`. Those regions become the atoms, and the levels above them
are **padded** (`MD0`, `MD00`):

```{r nonuts, eval = HAS_GEO}
lt <- as.data.frame(geoscales::geoscale_leaftable(gs))

# Every aggregation below is "roll the atoms up to a coarser geoframe", which is
# a lookup from the NUTS3 code to its parent at that level. Named once here
# rather than rebuilt inline each time.
parent_of <- function(level, from = "nuts3") setNames(lt[[level]], lt[[from]])

lt |>
  filter(nuts0 %in% c("BA", "MD", "UA", "XK")) |>
  group_by(nuts0) |>
  summarise(atoms = n(), nuts1 = first(nuts1), nuts2 = first(nuts2))
```

PyPSA-Eur repeats the adm1 code at every level, which makes each such region
its own parent. The derived hierarchy then contains pairs like `(MD-BA, MD-BA)`,
and energyRt's spatial roll-up becomes
`vOutTot[MD-BA] = ... + vOutTot[MD-BA]`, forcing every other term to zero
without error.

Eurostat's convention for identical geography is a distinct code per level:
Luxembourg is `LU → LU0 → LU00 → LU000`, and no NUTS country produces an
identity pair. The padding here follows that convention. "NUTS1" and "NUTS2" for
these four countries remain stand-ins rather than real administrative levels.

```{r nesting, eval = HAS_GEO}
f <- geoscales::geoscale_geoframes(gs)
vapply(seq_len(length(f) - 1L),
       function(i) isTRUE(geoscales::geoscale_nests(gs, f[i], f[i + 1L])),
       logical(1))
```

Every adjacent pair nests. Where nesting fails, a region has two parents and
its balance is added into both.

## 2. What each region carries

The leaftable holds 14 data columns per region, in two families. The family
determines what a column can be used for.

```{r columns, eval = HAS_GEO}
setdiff(names(lt), c("region", geoscales::geoscale_geoframes(gs)))
```

**Geographic**, read from the region shapes and defined for all 1,477 atoms:

| column | unit | kind |
|---|---|---|
| `km2` | km², equal-area projection | extensive |
| `pop` | thousands | extensive |
| `gdp` | EUR per capita | **intensive** |
| `gdp_total` | thousand EUR | extensive |

**Model**, read from the NUTS3 network and located at substation buses:

| column | unit | kind |
|---|---|---|
| `load_twh` | TWh per year | extensive |
| `peak_mw` | MW, coincident | **neither** |
| `n_buses` | substations in the region | extensive |
| `cap_mw`, `hydro_mw` | MW built today | extensive |
| `pot_onwind`, `pot_solar`, `pot_solar_hsat` | MW of potential | extensive |
| `pot_offwind_ac`, `pot_offwind_dc`, `pot_offwind_float` | MW of potential | extensive |

### Weights

```{r weights, eval = HAS_GEO}
geoscales::geoscale_weights(gs)
```

A weight splits a coarse quantity across atoms and averages an intensive one
back up. Seven columns are declared as weights, and the choice changes the
result. German demand split by area differs from the same demand split by
population or by measured demand:

```{r split, eval = HAS_GEO}
de <- filter(lt, nuts0 == "DE")

vapply(c("km2", "pop", "load_twh", "pot_onwind"),
       function(w) round(max(de[[w]]) / sum(de[[w]]), 4), 0)
```

The largest German region takes 1.5% of the country by area and 4.4% by
population. No check rejects an unsuitable weight, so the choice is the caller's
to make.

### `gdp` is intensive

```{r gdp, eval = HAS_GEO}
lt |>
  group_by(nuts0) |>
  summarise(plain_sum = round(sum(gdp)),
            pop_weighted_mean = round(sum(gdp * pop) / sum(pop)),
            gdp_total_bn = round(sum(gdp_total) * 1e3 / 1e9)) |>
  arrange(desc(gdp_total_bn)) |>
  head(5)
```

Summing `gdp` over Germany gives 15.6 million, which is not a quantity. The
population-weighted mean, ~41,800 EUR, is its GDP per capita. `gdp_total` is
shipped to provide an extensive money weight and gives ~3,475 bn EUR.

### The renewable potentials overlap

```{r potentials, eval = HAS_GEO}
lt |>
  group_by(nuts0) |>
  summarise(across(c(pot_onwind, pot_solar, pot_solar_hsat,
                     pot_offwind_ac, pot_offwind_dc, pot_offwind_float),
                   \(x) round(sum(x) / 1e3))) |>
  arrange(desc(pot_onwind)) |>
  head(5)
```

These columns must not be summed. `pot_solar` and `pot_solar_hsat` are
fixed-tilt and single-axis-tracking photovoltaics on the same land, and 313 of
the 314 offshore regions carry more than one `pot_offwind_*` type over the same
sea area. They ship as separate columns so no summing rule is imposed. The three
offshore columns are excluded from the weight set because they are zero for
every landlocked country.

### Where the model data is zero

```{r busless, eval = HAS_GEO}
lt |>
  summarise(regions = n(),
            no_substation = sum(n_buses == 0),
            pct = round(100 * mean(n_buses == 0), 1))
```

PyPSA-Eur places demand, plant and potential at substation buses, and 442
NUTS3 regions contain none. Every model column is zero there; the geographic
columns are unaffected. The zero reflects the model's allocation, in which a
busless region's land and load were folded into the neighbour it clusters with,
rather than the region itself.

The same applies at coarser levels:

```{r busless-up, eval = HAS_GEO}
bind_rows(lapply(c("nuts2", "nuts1", "nuts0"), function(l) {
  lt |>
    group_by(group = .data[[l]]) |>
    summarise(buses = sum(n_buses), .groups = "drop") |>
    summarise(level = l, groups = n(), no_substation = sum(buses == 0))
}))
```

Seven NUTS2 groups and three NUTS1 groups contain no substation. These are the
regions lost when 296 NUTS2 regions become 289 model nodes and 109 NUTS1 become
106. A weight summing to zero over a group divides by zero when averaging into
it; the affected groups are recorded on the object rather than substituted:

```{r degeneracy, eval = HAS_GEO}
names(attr(gs, "weight_degeneracy"))
```

Two entries are degenerate for a different reason. `GBI3` (Inner London) has no
generation and no land the availability rules admit for wind; `NO07`
(Nord-Norge) has no solar potential, lying above the Arctic Circle.

## 3. Load

```{r load-head}
head(nuts_load, 3)
```

### Energy is extensive

Going from 1,477 regions to 36 requires only a sum:

```{r load-agg, eval = HAS_GEO, fold = TRUE, collapse = FALSE}
fine <- filter(nuts_load, level == "nuts3")

agg_load <- function(col) {
  fine |>
    mutate(region = parent_of(col)[region]) |>
    group_by(region) |>
    summarise(load_twh = sum(load_twh), .groups = "drop") |>
    mutate(level = col)
}
derived <- bind_rows(lapply(c("nuts0", "nuts1", "nuts2"), agg_load))

# ...and it must agree with what was computed independently at each level.
nuts_load |>
  filter(level %in% c("nuts0", "nuts1", "nuts2")) |>
  select(level, region, load_twh) |>
  inner_join(derived, by = c("level", "region"),
             suffix = c("_shipped", "_derived")) |>
  group_by(level) |>
  summarise(regions = n(),
            max_abs_diff = max(abs(load_twh_derived - load_twh_shipped)))
```

The difference is zero at every level: the parts add to the whole, and no
weight is involved.

### Peak is not extensive

```{r peak, eval = HAS_GEO, fold = TRUE, collapse = FALSE}
pk <- function(fine_lv, coarse_lv) {
  f <- filter(nuts_load, level == fine_lv)
  m <- parent_of(coarse_lv, from = fine_lv)
  tibble(level = coarse_lv,
         summed = sum(tapply(f$peak_gw, m[f$region], sum)),
         coincident = sum(nuts_load$peak_gw[nuts_load$level == coarse_lv]))
}

bind_rows(
  pk("nuts3", "nuts2"), pk("nuts2", "nuts1"), pk("nuts1", "nuts0"),
  tibble(level = "europe",
         summed = sum(nuts_load$peak_gw[nuts_load$level == "nuts0"]),
         coincident = nuts_load$peak_gw[nuts_load$level == "europe"])) |>
  mutate(overstated_pct = round(100 * (summed / coincident - 1), 2))
```

In general `peak(parent) <= sum(peak(children))`, the gap being load
diversity. Here the gap is zero within every country and appears only across
Europe, at 6.8% (580 GW summed against 543 GW coincident).

This follows from the data rather than from European demand. PyPSA-Eur builds
per-bus load as a static fraction times the national hourly series, so every bus
in a country has the same normalised shape and their peaks coincide. Diversity
appears only where the shapes differ, which is between countries.

Resolution finer than a country therefore adds spatial detail to the level of
demand but not to its shape: at NUTS3, 1,477 regions share 36 distinct
profiles.

### 442 regions carry no demand

```{r coverage, eval = HAS_GEO}
nuts_load |>
  filter(level == "nuts3") |>
  summarise(regions = n(),
            no_demand = sum(n_buses == 0),
            pct = round(100 * mean(n_buses == 0), 1))
```

`load.substation_only: true` places demand only on substation buses, and 442
NUTS3 regions contain none; Germany alone has 182. They ship as explicit zeros
rather than missing rows. A NUTS3 model therefore has demand in 70% of its
regions, which bounds the resolution that is useful.

### The maps

Each region is shaded by its share of its parent's demand: a country against
Europe, a NUTS3 district against the NUTS2 region containing it. Shares of the
European total would instead restate the size of each region, which does not
change with resolution.

```{r map-helper, eval = MAPS, fold = TRUE, fold_label = "Show the map helpers"}
library(ggplot2)

# geoscale_geometry() dissolves the atoms up to the requested geoframe and
# returns an sf keyed by it, already carrying the CRS -- so the only work left
# is to give that key column the common name the joins below use.
level_sf <- function(lv) {
  geoscales::geoscale_geometry(nuts_gs, geoframe = lv) |>
    rename_with(\(nm) ifelse(nm == lv, "region", nm))
}

# The geoframes are ordered coarse -> fine, so a level's parent is the one
# before it: nuts3 -> nuts2, and nuts0 -> the single `europe` root.
GF <- geoscales::geoscale_geoframes(nuts_gs)
parent_frame <- function(lv) GF[[which(GF == lv) - 1L]]

load_map <- function(lv) {
  shares <- nuts_load |>
    filter(level == lv) |>
    select(region, load_twh) |>
    left_join(distinct(lt, region = .data[[lv]],
                       parent = .data[[parent_frame(lv)]]), by = "region") |>
    group_by(parent) |>
    # A parent carrying no demand at all divides by zero -- those are the
    # busless groups counted above. They map as missing, not as 0/0.
    mutate(total = sum(load_twh),
           share = if_else(total > 0, 100 * load_twh / total, NA_real_)) |>
    ungroup()

  level_sf(lv) |>
    left_join(select(shares, region, load_twh, share), by = "region")
}
BOX <- list(xlim = c(-11, 32), ylim = c(34, 71))   # crop overseas territories
```

```{r load-maps, eval = MAPS, fig.height = 7, fold = TRUE}
# rbind.sf, not bind_rows: the data-frame route drops the CRS and coord_sf()
# then fails inside the projection step.
LEVELS <- c("nuts0", "nuts1", "nuts2", "nuts3")

maps <- do.call(rbind, lapply(LEVELS, function(l) {
  s <- load_map(l); s$level <- l; s
}))
maps$level <- factor(maps$level, levels = LEVELS)

ggplot(maps) +
  geom_sf(aes(fill = share), colour = NA) +
  facet_wrap(~level, ncol = 2) +
  scale_fill_viridis_c(option = "C", trans = "sqrt",
                       name = "% of parent\nregion") +
  coord_sf(xlim = BOX$xlim, ylim = BOX$ylim, expand = FALSE) +
  theme_void() +
  theme(strip.text = element_text(face = "bold"))
```

At NUTS0 each country's share of Europe is at most 15%. At NUTS3 the shares
spread across the full range: most districts carry a small part of their NUTS2
region and a few carry nearly all of it. The dark patches are the 442 regions
with no substation, which carry none. Grey regions are those whose parent has no
demand at all — the seven NUTS2 groups counted above.

## 4. Transmission

`nuts_lines` holds every `base_s` line whose ends fall in different NUTS3
regions. The rest were dropped:

```{r lines-drop}
p <- attr(nuts_lines, "reneuro_provenance")
c(lines_total = p$lines_total, intra_nuts3_dropped = p$lines_intra_dropped,
  shipped = nrow(nuts_lines))
```

57% of lines never leave a NUTS3 region, and they carry 53% of the network's
capacity. The finest level available discards more than half the grid.

### Aggregating a corridor

```{r line-agg, eval = MAPS, fold = TRUE, fold_label = "Show the aggregation code"}
centroids <- function(level) {
  s <- level_sf(level)
  cc <- suppressWarnings(sf::st_coordinates(sf::st_centroid(sf::st_geometry(s))))
  tibble(region = s$region, cx = cc[, 1], cy = cc[, 2])
}

haversine <- function(x1, y1, x2, y2) {
  R <- 6371; r <- pi / 180
  a <- sin((y2 - y1) * r / 2)^2 +
       cos(y1 * r) * cos(y2 * r) * sin((x2 - x1) * r / 2)^2
  2 * R * asin(pmin(1, sqrt(a)))
}

agg_lines <- function(level, length_factor = 1.25) {
  m  <- parent_of(level)
  ct <- centroids(level)
  nuts_lines |>
    mutate(f = m[from], t = m[to]) |>
    filter(f != t) |>
    # canonical order, so A-B and B-A are one corridor
    mutate(a = pmin(f, t), b = pmax(f, t)) |>
    left_join(rename(ct, a = region, cx0 = cx, cy0 = cy), by = "a") |>
    left_join(rename(ct, b = region, cx1 = cx, cy1 = cy), by = "b") |>
    mutate(new_len = haversine(cx0, cy0, cx1, cy1) * length_factor,
           lf = if_else(length_km > 0, new_len / length_km, new_len)) |>
    group_by(from = a, to = b) |>
    summarise(s_nom = sum(s_nom), n_lines = n(),
              length_km = first(new_len),
              x = 1 / sum(1 / (lf * x)),        # parallel, length-rescaled
              cx0 = first(cx0), cy0 = first(cy0),
              cx1 = first(cx1), cy1 = first(cy1),
              .groups = "drop")
}
```

Three rules, taken from `pypsa.clustering.spatial`:

* `s_nom` **sums** — thermal ratings of parallel circuits add;
* `length` is **recomputed** as the great-circle distance between the two region
  centroids times `line_length_factor` (1.25) — it is *not* inherited;
* `x` is rescaled by the length ratio, then **parallel-combined** as
  `1 / Σ(1/x)`.

```{r line-table, eval = MAPS}
bind_rows(lapply(LEVELS, function(l) {
  a <- agg_lines(l)
  tibble(level = l, corridors = nrow(a),
         GW = round(sum(a$s_nom) / 1e3),
         median_km = round(median(a$length_km)),
         median_x = round(median(a$x), 2))
}))
```

> Checked against PyPSA-Eur's own 41-cluster network, built by its
> `cluster_network` rule: the capacity crossing a country boundary sums to
> 356,352.715 MW there and 356,352.715 MW in the table above. The corridor
> counts differ (77 against 65) because 41 clusters is not 36 countries. The
> comparison requires a clone and so is recorded here rather than run.

Note the length column. Aggregating to countries lengthens the network rather
than shortening it: a corridor becomes the distance between two national
centroids, which for DE–FR spans most of both countries. Impedance follows
length, so `r` rises with it, and since `loss ∝ r · s_nom²`, losses are
overestimated at coarse levels.

Capacity is overestimated for a separate reason: summing thermal ratings ignores
N-1 security and the loop flows that consume real boundary capacity. PyPSA-Eur
compensates with a flat `s_max_pu = 0.7`, applied identically at 4,382 nodes and
at 41.

The two errors act in opposite directions and do not cancel.

### The network as the model sees it

Straight chords between region centroids, which is what the aggregated model
contains:

```{r net-maps, eval = MAPS, fig.height = 7, fold = TRUE}
nets <- bind_rows(lapply(LEVELS, function(l) mutate(agg_lines(l), level = l)))
nets$level <- factor(nets$level, levels = LEVELS)

# white lines alone drown on plasma's bright end -- case them in dark grey
nets$lw <- scales::rescale(nets$s_nom / 1e3, to = c(0.1, 1.4))

ggplot() +
  geom_sf(data = maps, aes(fill = share), colour = NA) +
  geom_segment(data = nets,
               aes(x = cx0, y = cy0, xend = cx1, yend = cy1,
                   linewidth = I(lw + 0.5)),
               colour = "grey15", alpha = 0.85, lineend = "round") +
  geom_segment(data = nets,
               aes(x = cx0, y = cy0, xend = cx1, yend = cy1,
                   linewidth = s_nom / 1e3),
               colour = "white", lineend = "round") +
  facet_wrap(~level, ncol = 2) +
  scale_fill_viridis_c(option = "C", trans = "sqrt", name = "% of parent") +
  scale_linewidth_continuous(range = c(0.1, 1.4), name = "GW") +
  coord_sf(xlim = BOX$xlim, ylim = BOX$ylim, expand = FALSE) +
  theme_void() +
  theme(strip.text = element_text(face = "bold"),
        legend.key = element_rect(fill = "grey35", colour = NA))
```

At NUTS0 the mesh is reduced to 36 nodes and a small number of long chords. At
NUTS3 the structure of the European grid is recognisable, though half of it has
already been absorbed into the nodes.

## 5. Every quantity and its arithmetic

| quantity | kind | fine → coarse | coarse → fine |
|---|---|---|---|
| `km2`, `pop`, `gdp_total` | extensive | `sum` | split by weight |
| `load_twh`, `cap_mw`, `hydro_mw` | extensive | `sum` | split by weight |
| `pot_*` | extensive | `sum` **within a carrier** | split by weight |
| `n_buses` | extensive | `sum` | **cannot** be split |
| `gdp` (per capita) | intensive | `weighted_mean` by `pop` | copy |
| `peak_mw` | neither | **cannot** be derived — recompute from hourly | — |
| `s_nom` | extensive | `sum`, intra-region dropped | — |
| `length` | geometric | **recomputed** from centroids × 1.25 | — |
| `x`, `r` | reciprocal | rescale by length, then `1/Σ(1/x)` | — |
| load *shape* | — | already national; finer adds nothing | — |

The kind determines the reduction. Applying the wrong one produces no error: a
sum of per-capita GDP, of peaks, or across two overlapping potentials returns a
plausible number that does not correspond to any quantity.

## See also

* `vignette("reneuro")` — a model end to end, and restricting one to a region
  subset
* `vignette("about")` — how each model was built, solver benchmarks, references
* `?nuts_gs`, `?nuts_load`, `?nuts_lines`
