---
title: "Getting started with the model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with the model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
HAS_DPLYR <- requireNamespace("dplyr", quietly = TRUE)
HAS_GEO   <- requireNamespace("geoscales", quietly = TRUE) && HAS_DPLYR
```

```{r setup, message = FALSE}
library(energyRt)
library(reneuro)
```

This article uses only shipped data. It needs no PyPSA-Eur clone and no Python.
The results shown come from `be_solved`, a scenario solved in advance, so a
solver is not required either.

It covers four steps — pick a model, interpolate it, solve it, read the result —
and then restricting a model to a subset of its regions.

## 1. Pick a model

The models range from five nodes to a thousand, all of the same European
system. `vignette("about")` gives the resolutions and how each was built.

```{r models}
summary(pypsa_eur_5)
```

`pypsa_eur_5` covers five Belgian regions over one week: eleven generation
technologies, five supplies, two storages, five trade corridors and six weather
profiles holding the wind and solar availability.

`getObject()` returns the objects of one class, `get_region()` the declared
regions:

```{r objects}
names(getObject(pypsa_eur_5, class = "technology"))
get_region(pypsa_eur_5)
```

Each model carries its provenance as an attribute:

```{r provenance}
str(attr(pypsa_eur_5, "reneuro_provenance")[c("commit", "source", "built_on")])
```

## 2. Interpolate

`interpolate_model()` expands the model's declarations into the parameter
tables a solver needs: wildcards over regions, years and timeslices are
expanded, and the maps connecting processes to commodities are applied.

```{r interp, eval = FALSE}
scen <- interpolate_model(pypsa_eur_5, name = "be")
```

At continental scale this step dominates the runtime. The 1,035-node model
interpolates to roughly ten million parameter rows.

## 3. Solve

```{r solve, eval = FALSE}
scen <- write_script(scen, solver = solver_options$glpk)
scen <- read_solution(solve_scenario(scen, wait = TRUE))
```

GLPK is sufficient at this size. Past a few dozen regions, use Julia/HiGHS or
Pyomo/HiGHS; `vignette("about")` has the measured times, and
[`benchmark_solvers()`](../reference/benchmark_solvers.html) repeats them on
another machine.

Those two chunks produced `be_solved`, which ships with the package:

```{r objective}
# print() would round this to 9 significant digits; the gate is exact.
format(getData(be_solved, "vObjective", merge = TRUE)$value, digits = 16)
```

The build script refuses to save a scenario whose objective differs from that
figure, so it also serves as a regression test on the conversion.

## 4. Read the result

`getData()` reads variables, parameters and bounds alike. With `merge = TRUE`
it returns a long data frame with the index columns resolved to names.

```{r mix, eval = HAS_DPLYR, message = FALSE}
library(dplyr)

getData(be_solved, "vTechAct", merge = TRUE, process = TRUE,
        drop.zeros = FALSE, timeframe = "lowest") |>
  group_by(process) |>
  summarise(GWh = round(sum(value) / 1e3, 1)) |>
  arrange(desc(GWh)) |>
  head(8)
```

Gas and nuclear supply most of the week, followed by solar and offshore wind.
The same quantities are available from a PyPSA solve of the network, so the mix
can be compared directly.

## 5. Carve out a local model

`pypsa_eur_nuts3` is not intended to be solved at full resolution. Its use is
to supply a study area at NUTS3 granularity, which the coarser models cannot:
their aggregation has already averaged that detail away.

```{r subset, eval = FALSE}
pt <- geoscales::geoscale_leaftable(nuts_gs) |>
  as.data.frame() |>
  dplyr::filter(nuts0 == "PT") |>
  dplyr::pull(region) |>
  intersect(get_region(pypsa_eur_nuts3))

m <- attach_weather(pypsa_eur_nuts3, c("onwind", "solar", "ror")) |>
  subset_model_regions(region = pt)
```

The result is Portugal at 22 nodes — every NUTS3 region containing a
substation, with the network between them intact. `region` accepts any subset of
the model's regions, so a study area need not follow a national border.

`subset_model_regions()` drops the corridors crossing the boundary and reports
each. Without replacement the model is **islanded** and must meet demand from
its own resources, so its cost is an upper bound. `boundary_prices` replaces the
dropped routes with priced import/export stubs.

## Region data

`nuts_gs` is the region hierarchy the NUTS models are built from. It carries
geometry and per-region data: population, GDP, demand, existing capacity and
renewable potential.

```{r regions, eval = HAS_GEO}
geoscales::geoscale_leaftable(nuts_gs) |>
  as.data.frame() |>
  group_by(nuts0) |>
  summarise(twh = sum(load_twh), wind_gw = sum(pot_onwind) / 1e3) |>
  arrange(desc(twh)) |>
  head(5)
```

The column used to aggregate or split by is a modelling choice. National
demand divides by `load_twh` or `pop`, a wind target by `pot_onwind`; area is
rarely the right weight for either.

## Where next

* `vignette("data")` — the shipped datasets and their aggregation rules
* `vignette("about")` — how each model was built, solver benchmarks, references
* `?pypsa_eur_models`, `?nuts_gs` — the shipped objects and their licences
