---
title: "Getting started with the model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with the model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
HAS_DPLYR <- requireNamespace("dplyr", quietly = TRUE)
HAS_GEO   <- requireNamespace("geoscales", quietly = TRUE) && HAS_DPLYR
```

```{r setup, message = FALSE}
library(energyRt)
library(reneuro)
```

This article uses only shipped data. It needs no PyPSA-Eur clone and no Python.
The results shown come from `be_solved`, a scenario solved in advance, so a
solver is not required either.

It covers four steps — pick a model, interpolate it, solve it, read the result —
and then restricting a model to a subset of its regions.

## Settings

Four settings are worth knowing before the first solve. All are optional; the
defaults work.

**Where scenarios are written.** Every interpolation and solve writes files —
solver input, logs, the solution. By default they land under the working
directory; point that somewhere deliberate, especially if you keep model runs
outside the project:

```{r set-path, eval = FALSE}
set_scenarios_path("~/energy-models/scenarios")
get_scenarios_path()
```

**Which solver.** `solver_options` holds the ready-made back-end
configurations; setting a default saves passing `solver =` to every call:

```{r set-solver, eval = FALSE}
names(solver_options)                            # what is configured
set_default_solver(solver_options$julia_highs)   # or $pyomo_highs
```

HiGHS through Julia or Python is the one to set. GLPK will solve the five-node
models this article uses — and does, below — but it does not scale: at 41 nodes
it was still iterating after ten minutes on a problem the HiGHS barrier closes
in under four. `open_solvers()` reports what this machine has.

**How much it tells you.** Verbosity is a level, not a flag: `0` silent, `1`
progress, `2` and above increasingly detailed.

```{r set-verbose, eval = FALSE}
options(en.verbose = 1)     # or set_option("verbose", 1)
isVerbose()                 # TRUE at level >= 1
```

**A progress bar** for the long steps — interpolation of a continental model
takes minutes, and a bar is the difference between waiting and wondering:

```{r set-progress, eval = FALSE}
set_progress_bar(show = TRUE)
```

## 1. Pick a model

The models range from five nodes to a thousand, all of the same European
system. `vignette("about")` gives the resolutions and how each was built.

```{r models}
summary(pypsa_eur_5)
```

`pypsa_eur_5` covers five Belgian regions over one week: eleven generation
technologies, five supplies, two storages, five trade corridors and six weather
profiles holding the wind and solar availability.

`getObject()` returns the objects of one class, `get_region()` the declared
regions:

```{r objects}
names(getObject(pypsa_eur_5, class = "technology"))
get_region(pypsa_eur_5)
```

Each model carries its provenance as an attribute:

```{r provenance}
str(attr(pypsa_eur_5, "reneuro_provenance")[c("commit", "source", "built_on")])
```

## 2. Interpolate

`interpolate_model()` expands the model's declarations into the parameter
tables a solver needs: wildcards over regions, years and timeslices are
expanded, and the maps connecting processes to commodities are applied.

```{r interp, eval = FALSE}
scen <- interpolate_model(pypsa_eur_5, name = "be")
```

At continental scale this step dominates the runtime. The 1,035-node model
interpolates to roughly ten million parameter rows.

## 3. Solve

```{r solve, eval = FALSE}
scen <- write_script(scen, solver = solver_options$glpk)
scen <- read_solution(solve_scenario(scen, wait = TRUE))
```

GLPK is named here because it is what produced `be_solved`, and at five nodes
it is the quickest way to a result. It is the exception rather than the habit:
use Julia/HiGHS or Pyomo/HiGHS for anything larger. `vignette("about")` has the
measured times -- 10 s against 24 s at five nodes, 2,158 s against 79 s at
sixty-nine -- and
[`benchmark_solvers()`](../reference/benchmark_solvers.html) repeats them on
another machine.

Those two chunks produced `be_solved`, which ships with the package:

```{r objective}
# print() would round this to 9 significant digits; the gate is exact.
format(getData(be_solved, "vObjective", merge = TRUE)$value, digits = 16)
```

The build script refuses to save a scenario whose objective differs from that
figure, so it also serves as a regression test on the conversion.

## 4. Read the result

`getData()` reads variables, parameters and bounds alike. With `merge = TRUE`
it returns a long data frame with the index columns resolved to names.

```{r mix, eval = HAS_DPLYR, message = FALSE}
library(dplyr)

getData(be_solved, "vTechAct", merge = TRUE, process = TRUE,
        drop.zeros = FALSE, timeframe = "lowest") |>
  group_by(process) |>
  summarise(GWh = round(sum(value) / 1e3, 1)) |>
  arrange(desc(GWh)) |>
  head(8)
```

Gas and nuclear supply most of the week, followed by solar and offshore wind.
The same quantities are available from a PyPSA solve of the network, so the mix
can be compared directly.

## 5. Carve out a local model

The NUTS3 models are not intended to be solved at full resolution. Their use is
to supply a study area at NUTS3 granularity, which the coarser models cannot:
their aggregation has already averaged that detail away.

```{r subset, eval = FALSE}
pt <- geoscales::geoscale_leaftable(nuts_gs) |>
  as.data.frame() |>
  dplyr::filter(nuts0 == "PT") |>
  dplyr::pull(region) |>
  intersect(get_region(pypsa_eur_nuts3))

m <- attach_weather(pypsa_eur_nuts3, c("onwind", "solar", "ror")) |>
  subset_model_regions(region = pt)
```

`pypsa_eur_nuts3` ships without its weather profiles, so `attach_weather()`
puts back the ones the study needs.

The result is Portugal at 22 nodes — every NUTS3 region containing a
substation, with the network between them intact. `region` accepts any subset of
the model's regions, so a study area need not follow a national border.

`subset_model_regions()` drops the corridors crossing the boundary and reports
each. Without replacement the model is **islanded** and must meet demand from
its own resources, so its cost is an upper bound. `boundary_prices` replaces the
dropped routes with priced import/export stubs.

## Region data

`nuts_gs` is the region hierarchy the NUTS models are built from. It carries
geometry and per-region data: population, GDP, demand, existing capacity and
renewable potential.

```{r regions, eval = HAS_GEO}
geoscales::geoscale_leaftable(nuts_gs) |>
  as.data.frame() |>
  group_by(nuts0) |>
  summarise(twh = sum(load_twh), wind_gw = sum(pot_onwind) / 1e3) |>
  arrange(desc(twh)) |>
  head(5)
```

The column used to aggregate or split by is a modelling choice. National
demand divides by `load_twh` or `pop`, a wind target by `pot_onwind`; area is
rarely the right weight for either.

## Coarser models on demand

The NUTS0 and NUTS1 models are not shipped: they are an aggregation of the
NUTS3 one, so shipping them would store the same system twice more.

```{r aggregate, eval = FALSE}
m <- aggregate_model_regions(pypsa_eur_nuts3, level = "nuts1")
```

Capacities and demand are summed; efficiencies, availability factors and costs
take a mean weighted by the object's capacity in each region, so they stay
inside the range of the values they came from. Corridors that fall inside one
region are dropped and the rest merge. It gives 36 regions at `"nuts0"`, 106 at
`"nuts1"` and 289 at `"nuts2"`, in a few seconds.

## Where next

* `vignette("data")` — the shipped datasets and their aggregation rules
* `vignette("about")` — how each model was built, solver benchmarks, references
* `?pypsa_eur_models`, `?nuts_gs` — the shipped objects and their licences
