---
title: "Translation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Translation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
HAS_DPLYR <- requireNamespace("dplyr", quietly = TRUE)
```

```{r setup, message = FALSE}
library(reneuro)
library(energyRt)
```

This article documents how a PyPSA network becomes an energyRt model: which
component becomes which object, where a slot's value comes from, and the four
places where the two representations do not correspond one to one.

It describes the conversion behind the shipped models. Reading it is not
required to use them — `vignette("reneuro")` covers that — but it is what you
need to check a converted number against its source, or to decide whether a
question the model answers is one the translation supports.

Scope is the electricity-only network at the `base_s_{clusters}_elec.nc` cut
point: after the network is built, before policy overlays and temporal
averaging are applied.

## The component table

```{r map-table, echo = FALSE}
knitr::kable(data.frame(
  PyPSA = c("`Bus` carrier", "`Load`", "`Generator`", "`Generator` `p_max_pu` series",
            "`StorageUnit`", "`Store` + two `Link`s", "`Line` (AC)", "`Link` (DC)",
            "`StorageUnit` `inflow`"),
  energyRt = c("`commodity`", "`demand`", "`technology` (+ `supply`)", "`weather`",
               "`storage`", "`storage`", "`trade`", "`trade`",
               "`commodity` + `supply` + `storage` + `technology`"),
  Note = c("bus carriers only — see below",
           "series in `@demand`, one row per region and timeslice",
           "one object per carrier, regions in a `region` column",
           "kept out of the technology object",
           "one object; `duration` from `max_hours`",
           "one object with three part capacities",
           "transport, not Kirchhoff — see below",
           "transport, faithful",
           "four objects — see below")),
  col.names = c("PyPSA", "energyRt", ""))
```

Everything below is a case where that one-line summary is not the whole story.

## `carrier` names two different things

PyPSA's `carrier` field is used for two purposes, and only one of them is a
commodity:

* on a **bus** it names the energy flowing there — `AC`, `H2`, `battery` — and
  that is a commodity;
* on a **generator** or **link** it names the technology or the fuel — `CCGT`,
  `onwind`, `battery charger` — and that is not.

So commodities are taken from bus carriers, and generator carriers become
technology names. The `carriers` component itself is a lookup table of
attributes, not a set of objects to create.

## Generators carry their fuel explicitly

In an electricity-only network a generator has no fuel bus: `marginal_cost`
already folds fuel price divided by efficiency into one number. Translating
that literally gives an output-only technology — a CCGT that consumes no gas.
It is arithmetically equivalent, but it cannot carry a fuel price, a resource
limit, or an emission factor.

Thermal carriers therefore get an explicit fuel `@input` with `combustion = 1`
and the efficiency in `@ceff$cinp2use`. CO₂ then follows from the fuel's carbon
content on the commodity, rather than from a coefficient on the technology.

**The fuel supply is priced at zero and `varom` keeps the whole
`marginal_cost`.** The fuel price is not recoverable from the network file, and
charging the supply as well would count it twice. This is the one place in the
translation where double counting could enter.

Capacity depends on whether the generator is extendable:

| PyPSA | energyRt |
|---|---|
| not extendable | `@capacity$stock` — exogenous, no capital charged |
| extendable | `@capacity$cap.lo` / `cap.up` — built and paid for |

## Storage: two shapes in, one shape out

PyPSA has two storage representations and a network may use both:

* **`StorageUnit`** — one component carrying a power rating `p_nom`, a duration
  `max_hours`, and a round trip split into `efficiency_store` and
  `efficiency_dispatch`.
* **`Store` + two `Link`s** — an energy reservoir on its own bus, plus a charger
  and a discharger that carry the conversion efficiencies and their own ratings.

Both become **one** energyRt `storage`, because a storage has three independent
capacity variables — charge, discharge and reservoir — so the second shape needs
no extra objects:

```{r storage-shape, eval = HAS_DPLYR}
stg <- getObject(pypsa_eur_5, class = "storage")[["STG_BATTERY"]]
grep("^(inp|out|stg)\\.stock$", names(stg@capacity), value = TRUE)
stg@seff[, c("stgeff", "inpeff", "outeff")]
```

`inp.*` is the charger, `out.*` the discharger, `stg.*` the reservoir, and
`@seff` carries the two conversion efficiencies plus standing loss. A converted
model contains no separate charger or discharger technologies.

## Transmission is a transport model

PyPSA is a hybrid: Kirchhoff's voltage law applies to the AC `Line`s, which
form sub-networks and receive cycle constraints, while DC `Link`s are free
transport. energyRt's `trade` is transport throughout.

The consequence is asymmetric and worth stating plainly:

* **DC links translate faithfully** — a controllable branch is what a transport
  model represents natively.
* **AC lines are a relaxation.** The converted model's AC flows are less
  constrained than the source, so its objective is a **lower bound**, and it
  should not be used to draw transmission-expansion conclusions.

`energyRt::newACLine()` with `kvl = TRUE` at interpolation restores the cycle
constraints where that matters.

Capacity belongs to the trade object rather than to a route, so each bus pair
becomes its own object and parallel lines on a pair are merged by summing
`s_nom` — indistinguishable in a transport model in any case.

Quadratic AC losses are linearised. With `tranches = NULL` the corridor gets one
flat `teff` evaluated at a chosen loading; the default splits it into equal
capacity tranches with rising loss rates, which approximates the curve without
integer variables.

## Availability leaves the technology object

PyPSA keeps availability in two places at once. `p_max_pu` is a static column,
and `generators_t.p_max_pu` is a series covering the weather-driven generators
only. For every generator that has a series the static value is exactly 1, a
placeholder — so reading the static column alone would make every wind farm
permanently available.

energyRt keeps the bulk series out of the technology: a `weather` object holds
region × timeslice values and the technology references it by name.

```{r weather, eval = HAS_DPLYR}
names(getObject(pypsa_eur_5, class = "weather"))
```

## Hydro inflow becomes four objects

`storage_units_t.inflow` has no direct equivalent — an energyRt storage can be
charged from a commodity, but not from an exogenous series. A hydro reservoir
therefore decomposes:

| object | role |
|---|---|
| `commodity` `HYDWAT` | virtual reservoir energy, balanced hourly |
| `supply` `SUP_HYDWAT` | availability follows the inflow series |
| `storage` `STG_HYDRO` | the reservoir |
| `technology` `E_HYDRO` | `HYDWAT` → electricity at the dispatch efficiency |

## Conventions that differ

Two conventions produce correct-looking numbers that mean different things, and
both are places to check before comparing against a source model.

**Rating side.** A `Link`'s `p_nom` is rated on its input bus; an energyRt
capacity is rated on activity, i.e. output. Converting one to the other scales
by the efficiency: `capacity = p_nom * eff`, and `invcost = capital_cost / eff`.

**Cost annualisation.** Where `capital_cost` is already annuitised it maps to
`@invcost$eac`, not to `invcost` — supplying it as `invcost` would annuitise a
second time.

## What the translation does not do

The conversion reports rather than repairs. A parameter that cannot be
represented is named and its entity excluded; it is never replaced with a
plausible value. The register travels with each model:

```{r provenance}
str(attr(pypsa_eur_5, "reneuro_provenance")[c("describe", "commit", "source")])
```

## See also

* `vignette("reneuro")` — using a shipped model
* `vignette("about")` — how each model was built, and the licences
* `vignette("data")` — the region datasets and their aggregation rules
