---
title: "Data sources"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7.5, fig.height = 6, dpi = 96)
HAS_GEO <- requireNamespace("geoscales", quietly = TRUE) &&
           requireNamespace("sf", quietly = TRUE) &&
           requireNamespace("ggplot2", quietly = TRUE)
# `fold = TRUE` collapses a chunk's SOURCE behind a disclosure triangle and
# leaves its output alone, so the plumbing chunks stop interrupting the prose
# while their tables and maps stay on the page.
`%||%` <- function(x, y) if (is.null(x)) y else x
local({
  hook <- knitr::knit_hooks$get("source")
  knitr::knit_hooks$set(source = function(x, options) {
    out <- hook(x, options)
    if (!isTRUE(options$fold)) return(out)
    label <- options$fold_label %||% "Show code"
    paste0("<details>\n<summary>", label, "</summary>\n\n",
           out, "\n\n</details>\n")
  })
})
# precomputed figures: data-raw/data_sources_figures.R commits PNGs
# under figures/
fig <- function(name, dir = "data-sources")
  knitr::include_graphics(file.path("figures", dir, name))
```

```{r setup, message = FALSE}
library(reneuro)
library(dplyr)
```

The [41-node data page](data.html) shows the processed data as it enters
the model. This article is everything upstream of that: the raw sources,
how the renewable potentials and profiles are made, and the NUTS-level
datasets (`nuts_gs`, `nuts_load`, `nuts_lines`) with the aggregation
arithmetic that turns them into any coarser regional layer. The heavy
inputs (a 6.7 GB weather cutout, a PyPSA-Eur clone) do not ship, so their
figures are pre-rendered by `data-raw/data_sources_figures.R` and
committed; everything drawn from shipped data is computed live on this
page.

## 1. Weather: the cutout

Wind and solar availability come from one file: the *cutout*
`europe-2025-sarah3-era5.nc` — a 0.3° grid over Europe (177 × 131 cells) with
8,760 hourly fields of wind speed, solar irradiance, temperature and runoff
for 2025. This is *raw meteorology*, not capacity factors: 100 m wind speed
(ERA5, regridded from its native 0.25°), direct and diffuse solar influx
(SARAH-3 satellite data, aggregated from its native 0.05° and 30-minute
resolution). A 0.3° cell is roughly 21 × 33 km at European latitudes.
PyPSA-Eur turns it into per-region hourly capacity factors; those arrive in
the models as the `weather` objects (`W_ONWIND`, `W_SOLAR`, ...).

```{r cutout-wind, fig.cap="Mean 100 m wind speed over 2025, from the ERA5 fields of the cutout (CC-BY, Copernicus Climate Change Service)."}
fig("cutout-wind.png")
```

```{r cutout-solar, fig.cap="Mean solar influx (direct + diffuse) over 2025 — the SARAH-3 satellite record (CC-BY, EUMETSAT CM SAF) with ERA5 infill at high latitudes."}
fig("cutout-solar.png")
```

The exact licence notices ship in
`system.file("LICENSE.note", package = "reneuro")`.

## 2. Demand: measured ENTSO-E series

Hourly national load for 2025 comes from the ENTSO-E transparency platform
(with NESO for Great Britain). It lands in the models as each region's
`demand`, distributed below country level by a GDP/population proxy.

```{r demand-fig, fig.cap="Weekly-mean measured demand for the eight largest systems, ENTSO-E transparency platform. The workflow fills the few countries with gaps (UA, MK, CY, AL) from template years."}
fig("demand-entsoe.png")
```

The 2025 total across the modelled countries is 3,163 TWh — visible in the
shipped models as the sum over the demand objects:

```{r demand-total}
objs <- pypsa_eur_41@data[[1]]@data
dem <- objs[vapply(objs, function(o) class(o)[1], "") == "demand"]
sum(vapply(dem, function(d) sum(d@demand$demand), 0)) / 1e6  # TWh
```

## 3. The existing fleet

Power plants come from
[powerplantmatching](https://github.com/PyPSA/powerplantmatching), which
deduplicates GEM, JRC, ENTSO-E, GPD and other registries into one unit list —
157,549 units with coordinates and capacities. In the models they become the
non-extendable `stock` (nuclear, coal, hydro, ...) and the `cap.lo` floors on
extendable carriers (wind, solar); in the `pypsa_eur_41_<year>` horizon
series they also drive retirement, through announced closure dates where they
exist and assumed lifetimes elsewhere.

```{r fleet-fig, fig.cap="The existing fleet: units of 100 MW and above, sized by capacity, coloured by fuel — powerplantmatching (GEM, JRC, ENTSO-E, GPD and more, deduplicated)."}
fig("fleet.png")
```

## 4. The network source

Transmission comes from OpenStreetMap (which is what makes the shipped data
ODbL): substations and lines, simplified and clustered by PyPSA-Eur. In the
models it is the `trade` corridors — drawn on the
[41-node data page](data.html); the NUTS-level line dataset
and its aggregation arithmetic are documented below.

## 5. How potentials and profiles are made

Every wind and solar generator in the shipped models traces back to two
PyPSA-Eur artifacts: a **potential** (`p_nom_max`, the MW that may be built
per region) and a **profile** (`p_max_pu`, the hourly capacity factor). The
order of the pipeline matters more than any single step: **clustering comes
first.** PyPSA-Eur clusters a network that contains no generators at all,
and only then computes land eligibility, potentials, and profiles *directly
on the clustered regions* — the blending happens once, from weather cells
straight into whatever region set was chosen.

1. **Weather cutout** — the single 0.3° grid above.
2. **Land eligibility** — per carrier, a 100 m exclusion mask built from
   CORINE land cover, Natura 2000 protected areas, a 1 km urban buffer
   (onshore wind), ship density, water depth and shore distance (offshore).
   The eligible share of each 0.3° cell within each region becomes the
   *availability matrix*. **No criterion involves the resource itself** —
   no wind speed, no irradiation, no slope, no altitude.
3. **Potential** — `p_nom_max` = eligible area × a flat deployment density
   (3 MW/km² onshore wind, 5.1 solar, 2 offshore). Every eligible km²
   counts the same, whatever its wind.
4. **Profile** — capacity factors are computed per cell from the cutout
   (Vestas V112 3 MW power curve onshore; crystalline-silicon panel at 35°
   tilt), then blended into one regional series with weights proportional
   to each cell's *mean capacity factor* — "more generators are installed
   at better sites".

Two properties follow that most users don't expect:

**The asymmetry.** The potential counts *all* eligible land, but the profile
is weighted toward the *best* cells. A region therefore offers its entire
eligible area at close to its best-sites capacity factor — an optimistic
bias that grows with region size and internal resource spread. At 41 nodes
regions span whole countries; at NUTS3 the bias is mildest.

**Nothing is filtered by quality.** Regions whose mean wind capacity factor
is 0.2% survive alongside ones at 54%; a 0.094 MW potential is carried
through. A binning mechanism exists upstream (`resource_classes`) that would
split each region into capacity-factor tiers, but it is off by default and
off in these builds — each region gets exactly one blended generator per
carrier.

## 6. The land itself

The CORINE raster behind the exclusions, at CLC level 1, and what the
per-carrier grid-code inclusion admits from it. Wind's base is agriculture
*plus* forest and natural land; solar trades the forests for artificial
surfaces. About half the wind-only area is forest.

```{r corine-type, fig.cap="CORINE land cover at CLC level 1 (~5 km modal aggregate of the 250 m raster)."}
fig("corine_landtype.png", "data-sources")
```

```{r corine-elig, fig.cap="Land admitted by the PyPSA-Eur CORINE grid-code coding, per carrier. Inclusion only — Natura 2000, urban buffers and the other excluders apply on top."}
fig("corine_eligibility.png", "data-sources")
```

## 7. Available land

The availability matrices behind the shipped models, drawn at the cutout's
own resolution. Dark cells are excluded (urban fabric and its buffer for
wind, forests for solar, protected areas for both); the eligible remainder
is dominated by agricultural land, which no exclusion touches.

```{r avail-onwind, fig.cap="Available land for onshore wind: eligible share of each region-clipped cell after the land-use exclusions. The solid red line traces the CORINE raster coverage; outside it the shown fraction carries the flat 0.5 derate."}
fig("avail_map_onwind.png", "data-sources")
```

```{r mwmax-onwind, fig.cap="Installable onshore wind potential per cell: eligible area × 3 MW/km² (unfiltered; flat 0.5 derate outside the red CORINE coverage line)."}
fig("mwmax_map_onwind.png", "data-sources")
```

Reconciliation: eligible area × density reproduces PyPSA's `p_nom_max`
**exactly** (to the megawatt) for wind and all offshore carriers. Solar
differs by +1.8%, and the entire difference sits in cells north of 65°N:
SARAH-3 has no data there, and PyPSA silently drops that eligible land from
the solar potential rather than excluding it by policy.

## 8. The NUTS datasets

Three shipped datasets carry the regional layer the models are built on:
`nuts_gs` (the geography and the data attached to it), `nuts_load`
(demand) and `nuts_lines` (transmission). Each section below aggregates
one of them from NUTS3 to coarser levels and gives the arithmetic used,
which differs by quantity: capacities sum, per-capita values take a
weighted mean, and peaks and impedances have to be recomputed rather
than combined.

### Nested levels

```{r frames, eval = HAS_GEO}
gs <- nuts_gs
vapply(geoscales::geoscale_geoframes(gs),
       function(f) length(geoscales::geoscale_regions(gs, f)), 0L)
```

1,477 NUTS3 atoms group into 296 NUTS2, 109 NUTS1, 36 countries and a single
`europe` root. Every finer code has exactly one parent; the build asserts this
rather than assuming it, and the aggregation below depends on it.

The `europe` root is the only region at which a continental commodity balance
can be declared. `newCommodity(geoframe = "europe")` gives one Europe-wide CO2
allowance or gas market; without it the coarsest balance is per country, i.e. 36
separate markets.

```{r stack, eval = HAS_GEO, fig.height = 7, fig.cap = "The region hierarchy as a multilayer stack, each plane one level, coloured by each region's share of its parent's existing generation capacity (`cap_mw`)."}
library(geoscales)

lt <- as.data.frame(geoscale_leaftable(nuts_gs))

geoscale_autoplot(nuts_gs, type = "stack", direction = "down",
                  view = "oblique", colour = NA,
                  frame = T, frame_fill = ggplot2::alpha("grey60", 0.12),
                  data = lt, z = "cap_mw", rule = "share")

```

The same hierarchy as an icicle plot. The NUTS3 row is omitted: at 1,477
atoms most rectangles are thinner than their own outline, so the level reads
as a solid band and adds nothing:

```{r icicle, eval = HAS_GEO, fig.height = 3.4, fig.cap = "The region hierarchy as an icicle, `europe` down to NUTS2; widths are area shares."}
gs2 <- geoscales::geoscale_from_leaftable(
  lt |> summarise(km2 = sum(km2), .by = c(europe, nuts0, nuts1, nuts2)),
  geoframes = c("europe", "nuts0", "nuts1", "nuts2"),
  key = "nuts2", weights = "km2")
energyRt::plot_geoscale(gs2, type = "icicle")
```

Four countries — **BA, MD, UA, XK** — have no NUTS at all and are filled from
OpenStreetMap `adm1`. Those regions become the atoms, and the levels above them
are **padded** (`MD0`, `MD00`):

```{r nonuts, eval = HAS_GEO}
lt <- as.data.frame(geoscales::geoscale_leaftable(gs))

# Every aggregation below is "roll the atoms up to a coarser geoframe", which is
# a lookup from the NUTS3 code to its parent at that level. Named once here
# rather than rebuilt inline each time.
parent_of <- function(level, from = "nuts3") setNames(lt[[level]], lt[[from]])

lt |>
  filter(nuts0 %in% c("BA", "MD", "UA", "XK")) |>
  group_by(nuts0) |>
  summarise(atoms = n(), nuts1 = first(nuts1), nuts2 = first(nuts2))
```

PyPSA-Eur repeats the adm1 code at every level, which makes each such region
its own parent. The derived hierarchy then contains pairs like `(MD-BA, MD-BA)`,
and energyRt's spatial roll-up becomes
`vOutTot[MD-BA] = ... + vOutTot[MD-BA]`, forcing every other term to zero
without error.

Eurostat's convention for identical geography is a distinct code per level:
Luxembourg is `LU → LU0 → LU00 → LU000`, and no NUTS country produces an
identity pair. The padding here follows that convention. "NUTS1" and "NUTS2" for
these four countries remain stand-ins rather than real administrative levels.

```{r nesting, eval = HAS_GEO}
f <- geoscales::geoscale_geoframes(gs)
vapply(seq_len(length(f) - 1L),
       function(i) isTRUE(geoscales::geoscale_nests(gs, f[i], f[i + 1L])),
       logical(1))
```

Every adjacent pair nests. Where nesting fails, a region has two parents and
its balance is added into both.

### What each region carries

The leaftable holds 14 data columns per region, in two families. The family
determines what a column can be used for.

```{r columns, eval = HAS_GEO}
setdiff(names(lt), c("region", geoscales::geoscale_geoframes(gs)))
```

**Geographic**, read from the region shapes and defined for all 1,477 atoms:

| column | unit | kind |
|---|---|---|
| `km2` | km², equal-area projection | extensive |
| `pop` | thousands | extensive |
| `gdp` | EUR per capita | **intensive** |
| `gdp_total` | thousand EUR | extensive |

**Model**, read from the NUTS3 network and located at substation buses:

| column | unit | kind |
|---|---|---|
| `load_twh` | TWh per year | extensive |
| `peak_mw` | MW, coincident | **neither** |
| `n_buses` | substations in the region | extensive |
| `cap_mw`, `hydro_mw` | MW built today | extensive |
| `pot_onwind`, `pot_solar`, `pot_solar_hsat` | MW of potential | extensive |
| `pot_offwind_ac`, `pot_offwind_dc`, `pot_offwind_float` | MW of potential | extensive |

```{r weights, eval = HAS_GEO}
geoscales::geoscale_weights(gs)
```

A weight splits a coarse quantity across atoms and averages an intensive one
back up. Seven columns are declared as weights, and the choice changes the
result. German demand split by area differs from the same demand split by
population or by measured demand:

```{r split, eval = HAS_GEO}
de <- filter(lt, nuts0 == "DE")

vapply(c("km2", "pop", "load_twh", "pot_onwind"),
       function(w) round(max(de[[w]]) / sum(de[[w]]), 4), 0)
```

The largest German region takes 1.5% of the country by area and 4.4% by
population. No check rejects an unsuitable weight, so the choice is the caller's
to make.

`gdp` is intensive — summing it over Germany gives 15.6 million, which is
not a quantity; the population-weighted mean, ~41,800 EUR, is its GDP per
capita, and `gdp_total` ships as the extensive money weight:

```{r gdp, eval = HAS_GEO}
lt |>
  group_by(nuts0) |>
  summarise(plain_sum = round(sum(gdp)),
            pop_weighted_mean = round(sum(gdp * pop) / sum(pop)),
            gdp_total_bn = round(sum(gdp_total) * 1e3 / 1e9)) |>
  arrange(desc(gdp_total_bn)) |>
  head(5)
```

The renewable potentials overlap and must not be summed across carriers:
`pot_solar` and `pot_solar_hsat` are fixed-tilt and single-axis-tracking
photovoltaics on the same land, and 313 of the 314 offshore regions carry
more than one `pot_offwind_*` type over the same sea area. They ship as
separate columns so no summing rule is imposed; the offshore columns are
excluded from the weight set because they are zero for every landlocked
country.

```{r busless, eval = HAS_GEO}
lt |>
  summarise(regions = n(),
            no_substation = sum(n_buses == 0),
            pct = round(100 * mean(n_buses == 0), 1))
```

PyPSA-Eur places demand, plant and potential at substation buses, and 442
NUTS3 regions contain none. Every model column is zero there; the geographic
columns are unaffected. The same applies at coarser levels:

```{r busless-up, eval = HAS_GEO}
bind_rows(lapply(c("nuts2", "nuts1", "nuts0"), function(l) {
  lt |>
    group_by(group = .data[[l]]) |>
    summarise(buses = sum(n_buses), .groups = "drop") |>
    summarise(level = l, groups = n(), no_substation = sum(buses == 0))
}))
```

Seven NUTS2 groups and three NUTS1 groups contain no substation. A weight
summing to zero over a group divides by zero when averaging into it; the
affected groups are recorded on the object rather than substituted:

```{r degeneracy, eval = HAS_GEO}
names(attr(gs, "weight_degeneracy"))
```

### Load

```{r load-head}
head(nuts_load, 3)
```

Energy is extensive — going from 1,477 regions to 36 requires only a sum,
and it must agree with what was computed independently at each level:

```{r load-agg, eval = HAS_GEO, fold = TRUE, collapse = FALSE}
fine <- filter(nuts_load, level == "nuts3")

agg_load <- function(col) {
  fine |>
    mutate(region = parent_of(col)[region]) |>
    group_by(region) |>
    summarise(load_twh = sum(load_twh), .groups = "drop") |>
    mutate(level = col)
}
derived <- bind_rows(lapply(c("nuts0", "nuts1", "nuts2"), agg_load))

nuts_load |>
  filter(level %in% c("nuts0", "nuts1", "nuts2")) |>
  select(level, region, load_twh) |>
  inner_join(derived, by = c("level", "region"),
             suffix = c("_shipped", "_derived")) |>
  group_by(level) |>
  summarise(regions = n(),
            max_abs_diff = max(abs(load_twh_derived - load_twh_shipped)))
```

Peak is not extensive:

```{r peak, eval = HAS_GEO, fold = TRUE, collapse = FALSE}
pk <- function(fine_lv, coarse_lv) {
  f <- filter(nuts_load, level == fine_lv)
  m <- parent_of(coarse_lv, from = fine_lv)
  tibble(level = coarse_lv,
         summed = sum(tapply(f$peak_gw, m[f$region], sum)),
         coincident = sum(nuts_load$peak_gw[nuts_load$level == coarse_lv]))
}

bind_rows(
  pk("nuts3", "nuts2"), pk("nuts2", "nuts1"), pk("nuts1", "nuts0"),
  tibble(level = "europe",
         summed = sum(nuts_load$peak_gw[nuts_load$level == "nuts0"]),
         coincident = nuts_load$peak_gw[nuts_load$level == "europe"])) |>
  mutate(overstated_pct = round(100 * (summed / coincident - 1), 2))
```

In general `peak(parent) <= sum(peak(children))`, the gap being load
diversity. Here the gap is zero within every country and appears only across
Europe, at 6.8%: PyPSA-Eur builds per-bus load as a static fraction times the
national hourly series, so every bus in a country has the same normalised
shape and their peaks coincide. Resolution finer than a country therefore
adds spatial detail to the level of demand but not to its shape — at NUTS3,
1,477 regions share 36 distinct profiles. 442 regions carry no demand
(`load.substation_only: true` places demand only on substation buses).

```{r map-helper, eval = HAS_GEO, fold = TRUE, fold_label = "Show the map helpers"}
library(ggplot2)

level_sf <- function(lv) {
  geoscales::geoscale_geometry(nuts_gs, geoframe = lv) |>
    rename_with(\(nm) ifelse(nm == lv, "region", nm))
}

GF <- geoscales::geoscale_geoframes(nuts_gs)
parent_frame <- function(lv) GF[[which(GF == lv) - 1L]]

load_map <- function(lv) {
  shares <- nuts_load |>
    filter(level == lv) |>
    select(region, load_twh) |>
    left_join(distinct(lt, region = .data[[lv]],
                       parent = .data[[parent_frame(lv)]]), by = "region") |>
    group_by(parent) |>
    mutate(total = sum(load_twh),
           share = if_else(total > 0, 100 * load_twh / total, NA_real_)) |>
    ungroup()

  level_sf(lv) |>
    left_join(select(shares, region, load_twh, share), by = "region")
}
BOX <- list(xlim = c(-11, 41), ylim = c(34, 71))   # crop overseas territories
```

```{r load-maps, eval = HAS_GEO, fig.height = 6, fold = TRUE, fig.cap = "Each region's share of its parent's demand, per level. Dark = little of the parent's demand (including the 442 substation-less regions); grey = the parent itself carries no demand."}
LEVELS <- c("nuts0", "nuts1", "nuts2", "nuts3")

maps <- do.call(rbind, lapply(LEVELS, function(l) {
  s <- load_map(l); s$level <- l; s
}))
maps$level <- factor(maps$level, levels = LEVELS)

ggplot(maps) +
  geom_sf(aes(fill = share), colour = NA) +
  facet_wrap(~level, ncol = 2) +
  scale_fill_viridis_c(option = "C", trans = "sqrt",
                       name = "% of parent\nregion") +
  coord_sf(xlim = BOX$xlim, ylim = BOX$ylim, expand = FALSE) +
  theme_void() +
  theme(strip.text = element_text(face = "bold"))
```

### Transmission lines

`nuts_lines` holds every `base_s` line whose ends fall in different NUTS3
regions. The rest were dropped:

```{r lines-drop}
p <- attr(nuts_lines, "reneuro_provenance")
c(lines_total = p$lines_total, intra_nuts3_dropped = p$lines_intra_dropped,
  shipped = nrow(nuts_lines))
```

57% of lines never leave a NUTS3 region, and they carry 53% of the network's
capacity. The finest level available discards more than half the grid.

```{r line-agg, eval = HAS_GEO, fold = TRUE, fold_label = "Show the aggregation code"}
centroids <- function(level) {
  s <- level_sf(level)
  cc <- suppressWarnings(sf::st_coordinates(sf::st_centroid(sf::st_geometry(s))))
  tibble(region = s$region, cx = cc[, 1], cy = cc[, 2])
}

haversine <- function(x1, y1, x2, y2) {
  R <- 6371; r <- pi / 180
  a <- sin((y2 - y1) * r / 2)^2 +
       cos(y1 * r) * cos(y2 * r) * sin((x2 - x1) * r / 2)^2
  2 * R * asin(pmin(1, sqrt(a)))
}

agg_lines <- function(level, length_factor = 1.25) {
  m  <- parent_of(level)
  ct <- centroids(level)
  nuts_lines |>
    mutate(f = m[from], t = m[to]) |>
    filter(f != t) |>
    mutate(a = pmin(f, t), b = pmax(f, t)) |>
    left_join(rename(ct, a = region, cx0 = cx, cy0 = cy), by = "a") |>
    left_join(rename(ct, b = region, cx1 = cx, cy1 = cy), by = "b") |>
    mutate(new_len = haversine(cx0, cy0, cx1, cy1) * length_factor,
           lf = if_else(length_km > 0, new_len / length_km, new_len)) |>
    group_by(from = a, to = b) |>
    summarise(s_nom = sum(s_nom), n_lines = n(),
              length_km = first(new_len),
              x = 1 / sum(1 / (lf * x)),        # parallel, length-rescaled
              cx0 = first(cx0), cy0 = first(cy0),
              cx1 = first(cx1), cy1 = first(cy1),
              .groups = "drop")
}
```

Three rules, taken from `pypsa.clustering.spatial`:

* `s_nom` **sums** — thermal ratings of parallel circuits add;
* `length` is **recomputed** as the great-circle distance between the two region
  centroids times `line_length_factor` (1.25) — it is *not* inherited;
* `x` is rescaled by the length ratio, then **parallel-combined** as
  `1 / Σ(1/x)`.

```{r line-table, eval = HAS_GEO}
bind_rows(lapply(LEVELS, function(l) {
  a <- agg_lines(l)
  tibble(level = l, corridors = nrow(a),
         GW = round(sum(a$s_nom) / 1e3),
         median_km = round(median(a$length_km)),
         median_x = round(median(a$x), 2))
}))
```

> Checked against PyPSA-Eur's own 41-cluster network, built by its
> `cluster_network` rule: the capacity crossing a country boundary sums to
> 356,352.715 MW there and 356,352.715 MW in the table above. The corridor
> counts differ (77 against 65) because 41 clusters is not 36 countries. The
> comparison requires a clone and so is recorded here rather than run.

Aggregating to countries lengthens the network rather than shortening it: a
corridor becomes the distance between two national centroids, so `r` rises
and losses are overestimated at coarse levels. Capacity is overestimated for
a separate reason: summing thermal ratings ignores N-1 security and loop
flows. PyPSA-Eur compensates with a flat `s_max_pu = 0.7`; the two errors
act in opposite directions and do not cancel.

```{r net-maps, eval = HAS_GEO, fig.height = 7, fold = TRUE, fig.cap = "The NUTS-level transmission aggregated to each level: straight chords between region centroids, width proportional to summed thermal rating."}
nets <- bind_rows(lapply(LEVELS, function(l) mutate(agg_lines(l),
                                                    level = l)))
nets$level <- factor(nets$level, levels = LEVELS)

ggplot() +
  geom_sf(data = maps, fill = "wheat", colour = "white", linewidth = 0.1) +
  geom_segment(data = nets,
               aes(x = cx0, y = cy0, xend = cx1, yend = cy1,
                   linewidth = s_nom / 1e3),
               colour = "dodgerblue", alpha = 0.8, lineend = "round") +
  facet_wrap(~level, ncol = 2) +
  scale_linewidth_continuous(range = c(0.1, 1.4), name = "GW") +
  coord_sf(xlim = BOX$xlim, ylim = BOX$ylim, expand = FALSE) +
  theme_void() +
  theme(strip.text = element_text(face = "bold"))
```

### Every quantity and its arithmetic

| quantity | kind | fine → coarse | coarse → fine |
|---|---|---|---|
| `km2`, `pop`, `gdp_total` | extensive | `sum` | split by weight |
| `load_twh`, `cap_mw`, `hydro_mw` | extensive | `sum` | split by weight |
| `pot_*` | extensive | `sum` **within a carrier** | split by weight |
| `n_buses` | extensive | `sum` | **cannot** be split |
| `gdp` (per capita) | intensive | `weighted_mean` by `pop` | copy |
| `peak_mw` | neither | **cannot** be derived — recompute from hourly | — |
| `s_nom` | extensive | `sum`, intra-region dropped | — |
| `length` | geometric | **recomputed** from centroids × 1.25 | — |
| `x`, `r` | reciprocal | rescale by length, then `1/Σ(1/x)` | — |
| load *shape* | — | already national; finer adds nothing | — |

The kind determines the reduction. Applying the wrong one produces no error: a
sum of per-capita GDP, of peaks, or across two overlapping potentials returns a
plausible number that does not correspond to any quantity.

## 9. Known caveats

- **Not every upstream dataset covers the whole modeled territory.** Some
  of the redistributable datasets behind the exclusions are unavailable for
  parts of the PyPSA-Eur domain; where that is the case the affected
  exclusion is skipped upstream and availability is overstated. The
  improved layer draws the coverage boundary as a solid red line on the
  potential maps and applies a flat 0.5 availability derate outside it
  (`non_corine_derate` in `params.yml`) as a conservative stand-in.
  Extending the coverage with data from other sources is possible — see
  the source table below and the package's `LICENSE.note` for what is
  used and under which terms.
- **Capacity-factor granularity is ~25–30 km.** Eligibility is known at
  100 m, but the resource is only known per 0.3° cell — a windy ridge and a
  calm valley inside one cell are indistinguishable.

## 10. Sources at a glance

| source | licence | vintage | becomes |
|---|---|---|---|
| ERA5 (Copernicus C3S) | CC-BY | 2025 hourly | wind / hydro / temperature profiles → `weather` objects |
| SARAH-3 (EUMETSAT CM SAF) | CC-BY | 2025 hourly | solar profiles → `weather` objects |
| ENTSO-E transparency (+ NESO for GB) | open re-use | 2025 hourly | national load → `demand` |
| powerplantmatching | GPL-3 code, mixed open registries | through 2024 | fleet → `stock`, `cap.lo` floors, retirement in the horizon series |
| OpenStreetMap (+ TYNDP projects) | ODbL | 2026 extract | grid → `trade` corridors (the reason the data licence is ODbL) |
| NUTS / JRC / Eurostat layers | EU open licences | mixed | region geometry + attributes → `nuts_gs` |
| Natural Earth (via rnaturalearth) | public domain | 1:50m | coastlines and country contours on the figures |
| technology-data v0.14.0 | CC-BY | 2020–2050 in 5-year steps | costs → `invcost` / `varom` / supply prices |

What is *not* here: the World Database on Protected Areas (non-redistributable;
excluded, with the effect documented in `LICENSE.note`) and the OPSD demand
series (replaced by the 2025 ENTSO-E pull).
