---
title: "Solar energy"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE,
                      # figures are written at mixed dpi (150 and 200), which
                      # knitr turns into mixed display widths; every file is
                      # wider than the page column, so a uniform width only
                      # ever downscales
                      out.width = "100%")
# All figures are precomputed by data-raw/cf-workflow/ (the sources live only
# on the build machine) and committed under figures/cf-pipeline/.
fig <- function(name) knitr::include_graphics(file.path("figures/cf-pipeline", name))
```

The solar half of the quality-aware resource layer built in
`data-raw/cf-workflow/` (the wind half, with the atlas classification and the
calibration machinery, is in the [wind article](wind-potential.html); the
upstream pipeline is in the [data-sources article](data-sources.html)).

PyPSA-Eur's solar capacity factors are likewise untouched and ship with the
replica, taking both potential and variability from the 2025 cutout (SARAH-3
irradiance, ERA5 for the rest). The Global Solar Atlas (v2, ~1 km PVOUT)
provides the long-term benchmark; the cutout remains the source of hourly
variability. Solar is less diverse by location than wind: the resource is not
terrain-driven at the cutout's ~25 km, and the two instruments agree within
±0.02 across the 41 regions, where wind's gap reaches +0.19. The level
therefore needs no calibration and no low-quality filter. What the atlas adds
is a supply curve — six 0.02-wide quality levels across the continent's
0.09–0.20 range, so a cluster's land is offered at the level it occupies
rather than at one blended number. As on the wind side, this is not part of
the shipped models: it arrives with the repository update and ships
separately ([revised assumptions](translation.html#capacity-factors)).

## Available land

Solar's eligibility differs from wind's mainly in what it admits: urban
and artificial surfaces are *in* (rooftop and urban PV), forests are out.

```{r avail-solar, fig.cap="Available land for solar: eligible share of each region-clipped cell after the land-use exclusions, with the 0.5 coverage derate applied outside the solid red CORINE coverage line."}
fig("avail_map_solar.png")
```

One reconciliation caveat is worth knowing: eligible area × density
reproduces PyPSA's solar `p_nom_max` except for +1.8%, and the entire
difference sits in cells north of 65°N — SARAH-3 has no data there, and
PyPSA silently drops that eligible land from the solar potential rather
than excluding it by policy. Groups with no usable solar irradiance data
are likewise excluded from this layer.

## Clustering

Per-cell hourly capacity factors (the same fixed-tilt panel assumptions
as PyPSA) are clustered by profile shape — weighted k-medoids on
correlation distance over mean-normalized series, area weights — with the
number of clusters per country the smallest that keeps the weighted
information loss under the 5% tolerance. Solar profiles are dominated by
latitude and the shared diurnal cycle, so the whole continent resolves to
**62 clusters** (wind, at its own tolerance, needs 287):

```{r cluster-cf-solar, fig.cap="Solar clusters on the region-clipped cells (cell × NUTS1), coloured by mean capacity factor."}
fig("cluster_cf_map_solar.png")
```

What the model finally receives is neither of those. Its unit of supply is a
**location**: one (cluster × level) pair, holding its own piece of land at its
own capacity factor. The pairs do not overlap — within a country their areas
sum to their union — so a location is a real place, not an accounting
category, and a country offers as many as its clusters have levels. Spain is
a single cluster carved into six levels, so it offers six locations, not one;
France's four clusters carry sixteen between them.

```{r solar-supply-map, fig.cap="The solar layer as the model receives it: every installable piece coloured by its location number — the (cluster × level) pair's rank within its country, best first — with white cluster boundaries on top. One location is one colour inside one outline. Spain's single cluster carves into six locations, France's four into sixteen; the subtitle counts those that reach the nodes with capacity, matching the table below."}
fig("supply_map_solar.png")
```

## The solar atlas check

The wind layer earned its atlas treatment because the Global Wind Atlas
disagrees with the cutout systematically (median ratio 0.63). The solar
mirror of that comparison uses the [Global Solar
Atlas](https://globalsolaratlas.info) (World Bank/ESMAP, Solargis data,
CC-BY-4.0): PVOUT, the long-term average daily specific yield in
kWh/kWp at ~1 km, read as a capacity factor (`PVOUT / 24`) over the
same eligible land. Two assumption differences are worth stating before
the map: the atlas assumes optimally tilted crystalline-silicon PV with
its own performance-ratio chain (~0.81), while PyPSA/atlite uses a
fixed 35° tilt and its own loss model; and the atlas raster stops at
65°N — the same northern limit as SARAH-3, so the no-data groups
already excluded from this layer stay excluded.

```{r gsa-map, fig.cap="GSA PVOUT over the solar-eligible cells, read as a capacity factor (PVOUT / 24). Grey = no atlas data (≥ 65°N, the same limit as SARAH-3)."}
fig("gsa_cf_map.png")
```

```{r gsa-carved-cat, fig.cap="The same atlas carved into the six quality classes at ~1 km — all available atlas data in the map window, beyond the model domain (Turkey, the African rim, the eastern neighbours), with country outlines for orientation. The gradient is mostly latitude, but real sub-country structure survives: Alpine and coastal pockets, the Scottish Highlands below 0.10, Iberia above 0.18. Blank land carries no atlas data: north of 65°N, and Iceland, which the Solargis model does not cover. These classes are the levels the supply curve is built on."}
fig("gsa_carved_map_cat.png")
```

```{r gsa-hist, fig.cap="Where the eligible solar capacity sits on that scale: MW-weighted density over the GSA capacity factor, segmented by the same six classes. The whole continent spans ~0.09–0.20 — one wind class wide."}
fig("gsa_cf_hist.png")
```

The comparison itself is flat. Region by region the two instruments
agree to within ±0.02: median PyPSA 0.138 vs GSA 0.140 (median ratio
0.97), the whole 41-region spread inside [−0.005, +0.018] — against
wind's 0.63 ratio and gaps of ±0.19. The mild pattern that remains is
the assumption delta and satellite-product differences at high
latitudes (the Baltics read ~0.015 higher in the atlas), not terrain:

```{r gsa-scatter, fig.cap="Cell-level cutout CF vs GSA CF over eligible solar land: correlation 0.85, median ratio 0.96, no saturation or terrain branch. The red line is 1:1, not a fit — the cloud sits on it."}
fig("gsa_vs_cutout.png")
```

Conditioning the atlas side on each quality level — the wind article's
tier view, built the same way — makes the same point in the negative.
Where wind's best land runs +0.35 above its region's blend, solar's
best level barely separates from its lowest: the blend is a fair price
for a region's solar land at every quality level, which is why the
layer needs no calibration.

```{r cf-vs-gsa-tiers, fig.cap="The atlas-vs-ERA5 gap painted on the carved installable land: each piece coloured by the conditional Global Solar Atlas mean of its quality level minus the region's SARAH/ERA5 mean. The quality bands are spatially exclusive, so one map carries all levels (the open lowest band is omitted — the pass-fraction grid starts at 0.08). Note the scale: the whole range is a fraction of the wind analog's."}
fig("cf_vs_gsa_41_tiers.png")
```

## One shape set, several levels

The atlas check is why solar needs no *calibration* where wind did:
the two instruments already agree within the stated assumption delta,
and a 0.3° cell hides no ridge-and-valley structure — so the shape
clusters stand as they are. What the atlas does add, carved the same
way as the wind atlas (the carving machinery is raster-agnostic), is
**quality levels**: six 0.02-wide classes spanning the continent's
0.09–0.20 range, splitting each cluster's installable area into a
supply curve — exactly the wind tier pattern, one hourly shape per
cluster and per-level area on top.

Solar resolves to **62 shape clusters** carrying **194 locations**, against
wind's 287 and 1,003 — the same continent, a fifth
of the structure, because solar profiles differ far less by location.
Mapped onto the 41 nodes those become 227 supply rows, roughly five and a
half for every blended generator the replica carries. Capacity is close to
the original's, 12,161 GW against 13,893: no quality filter removes solar
land, so what separates them is the coverage derate and the atlas's own
gap north of 65°N.

```{r tbl-clusters-solar}
cc <- read.csv("tables/cluster_counts.csv")
d <- cc[cc$resource == "solar",
        c("country", "nodes", "clusters", "af_cl", "gw_pypsa", "gw")]
d <- rbind(d, data.frame(country = "Total", nodes = sum(d$nodes),
                         clusters = sum(d$clusters), af_cl = sum(d$af_cl),
                         gw_pypsa = sum(d$gw_pypsa), gw = sum(d$gw)))
d$gw <- round(d$gw); d$gw_pypsa <- round(d$gw_pypsa)
knitr::kable(
  d, row.names = FALSE, align = "lrrrrr", format.args = list(big.mark = ","),
  col.names = c("Country", "41-node regions", "Shape clusters",
                "Locations", "Original GW", "This layer GW"),
  caption = "Solar resource clusters by country: the 41-node regions the country holds, the shape clusters the layer resolves, and the locations those carve out — one (cluster × level) pair each, counted only where it carries capacity at the nodes. Original GW is PyPSA-Eur's own `p_nom_max` for the same land; this layer's GW is after the coverage derate.")
```

The same test applied to wind — does a solve take the better bands first? —
answers differently here, and in a way consistent with everything above.
Solved at 2050 with a zero carbon cap the model builds 1,217 GW of the 12,161
on offer, 10%, but the order is not the quality order: the 0.12–0.14 band
carries nearly half of all solar built while only an eighth of it is used,
and the 0.14–0.16 band above it is barely touched. Where wind's utilization
climbs monotonically with quality, solar's does not, because a 0.02-wide
level is worth less than proximity to demand and to the network.

```{r tbl-used-solar}
cu <- read.csv("tables/capacity_used_by_tier.csv")
d <- cu[cu$resource == "solar",
        c("tier_lab", "steps", "max_gw", "built_gw", "used_pct", "share_of_built")]
d$max_gw <- round(d$max_gw); d$built_gw <- round(d$built_gw)
d$used_pct <- round(d$used_pct, 1); d$share_of_built <- round(d$share_of_built, 1)
knitr::kable(
  d, row.names = FALSE, align = "lrrrrr", format.args = list(big.mark = ","),
  col.names = c("Level", "Locations", "Max GW", "Built GW",
                "% of level built", "% of build"),
  caption = "What the 2050 zero-carbon solve takes from each solar level: capacity on offer, capacity built, the share of the level's own limit taken, and the level's share of everything built. Unlike wind, utilization does not follow quality — siting is decided by other things.")
```

Of the 12.2 TW unfiltered solar potential, 8.0 TW sits in the two
middle classes (0.12–0.16), 2.1 TW above 0.16 (Iberia, the
Mediterranean rim), and only 0.1 TW below 0.10 (the Scottish Highlands
and the Norwegian coast). The stored decomposition
(`tier_areas_solar.parquet`) carries the 194 locations
over the 62 shape clusters. Unlike wind, no low-quality land is
filtered out — the flat-gap verdict above means even the lowest class
is measured correctly by the cutout; the levels differentiate the
supply curve rather than screen locations.
